---
title: "Alberta and western Canada site index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alberta and western Canada site index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CanWestSiteIndexCFS)
```

The British Columbia side of this package wraps the Sindex height-age curves.
This vignette covers the rest of western Canada: the Alberta GYPSY height-age
models, and the three ways of estimating site index when no suitable top height
tree is available.

## Which estimator should I use?

Bjelanovic and Comeau (2019) and Comeau (2020) give an explicit preference
order for a stand where site index cannot be measured directly:

1. **Edatope** -- `si_from_edatope()`. Preferred: edatopes cover a narrower
   range of site conditions than ecosites and are easier to determine in young
   stands.
2. **Ecosite** -- `si_from_ecosite()`. Same study, coarser stratification.
3. **Species conversion** -- `ab_si_to_si()`. Use only when a measured site
   index for another species is available, and compare the result against the
   ecological estimate.
4. **Provincial ecosite guides** -- `si_from_ecosite_guide()`. Retained for
   reference and for British Columbia and Manitoba, which the 2019 study does
   not cover. The Alberta guide values are based on mature and old growth
   sampling and superseded height-age equations; Comeau (2020) recommends
   `si_from_edatope()` in their place.

## Age basis: the thing to get right first

Alberta carries two site index definitions and they are not interchangeable.

- **SIt** -- top height at 50 years **total** age. This is what the GYPSY
  height-age models are parameterised on.
- **SIbh** -- top height at 50 years **breast height** age. This is what
  Alberta operational inventory and the Mixedwood Growth Model use.

Because a tree needs some years to reach 1.3 m, breast height age 50 is total
age `50 + Y2BH`, so `SIbh` is always the larger number.

```{r}
ab_si_total_to_bh(c("Aw", "Sb", "Pl", "Sw"), 20)
ab_years_to_bh(c("Aw", "Sb", "Pl", "Sw"), 20)
```

GYPSY derives years-to-breast-height from site index rather than treating it as
a species constant, so a poorer site takes longer:

```{r}
ab_years_to_bh("Pl", c(10, 15, 20, 25))
```

The reverse conversion has no closed form and is solved numerically:

```{r}
ab_si_bh_to_total("Pl", 21.3314)
```

## Height and site index

`ab_si_to_height()` evaluates the curve; `ab_height_to_si()` inverts it. Both
take **total** age.

```{r}
# At the reference age, top height equals SIt by construction
ab_si_to_height(c("Aw", "Sb", "Pl", "Sw"), 20, 50)

# A lodgepole pine height trajectory
data.frame(
  total_age  = seq(10, 100, by = 10),
  top_height = round(ab_si_to_height("Pl", 18, seq(10, 100, by = 10)), 2)
)
```

Going the other way from a field measurement, and asking for the result on the
basis MGM expects:

```{r}
ab_height_to_si("Aw", top_height = 17.5, total_age = 40)
ab_height_to_si("Aw", top_height = 17.5, total_age = 40, age_basis = "breast")
```

### Species beyond the four GYPSY curves

GYPSY covers aspen, black spruce, lodgepole pine and white spruce. Other
species are conventionally modelled with the most similar curve.
`ab_gypsy_species()` makes that substitution visible rather than silent:

```{r}
ab_gypsy_species(c("Aw", "Pj", "Bw", "Pb", "Fb", "Sw"))
```

An unrecognised code returns `NA` with a warning rather than a plausible
looking number:

```{r}
suppressWarnings(ab_si_to_height("Zz", 18, 40))
```

## Site index from edatope

Edatope is the combination of soil moisture regime (SMR, 2-8) and soil nutrient
regime (SNR, B/C/D) within a natural subregion.

```{r}
si_from_edatope("LF", smr = 5, snr = "C", species = c("Aw", "Sw", "Pl"))
```

Ask for `details = TRUE` when the sample size matters -- several cells rest on
only three or four plots:

```{r}
si_from_edatope("LF", 5, "C", c("Aw", "Sw", "Pl"), details = TRUE)
```

Combinations that were never sampled, or that had fewer than three plots, come
back as `NA` with a warning:

```{r}
suppressWarnings(si_from_edatope("CM", 9, "D", "Aw"))
```

## Site index from ecosite

```{r}
si_from_ecosite("CM", "d", c("Aw", "Sw", "Pj"))
si_from_ecosite("CM", "d", c("Aw", "Sw", "Pj"), age_basis = "total")
```

## Species site index conversion

Six directed pairs are published, on each of the two age bases:

```{r}
ab_si_conversions[ab_si_conversions$age_basis == "breast",
                  c("from", "to", "intercept", "slope", "see", "pearson_r")]
```

```{r}
ab_si_to_si("Aw", 20, "Sw")
ab_si_to_si("Aw", 20, "Sw", age_basis = "total")
```

Aw-Pl, Pl-Pj and Sw-Pj were **not** significant and are deliberately absent:

```{r}
suppressWarnings(ab_si_to_si("Aw", 20, "Pl"))
```

These equations explain only 26-47 % of the variation in predicted site index.
Comparing against the ecological estimate is the recommended check:

```{r}
aw <- si_from_edatope("CM", 5, "C", "Aw")
data.frame(
  source            = c("edatope", "converted from Aw"),
  sw_site_index = round(c(si_from_edatope("CM", 5, "C", "Sw"),
                          ab_si_to_si("Aw", aw, "Sw")), 2)
)
```

The reverse equations were fitted independently rather than derived
algebraically, so a conversion is not reversible:

```{r}
ab_si_to_si("Sw", ab_si_to_si("Aw", 20, "Sw"), "Aw")
```

## Provincial ecosite guides

`si_from_ecosite_guide()` covers British Columbia site series, the Alberta
ecosite guides, both Saskatchewan guides and the Manitoba guide.

```{r}
si_from_ecosite_guide("BWBSmw", "110", c("Pl", "Sw", "Sb", "Aw"))
si_from_ecosite_guide("Lower Foothills", "e", c("Aw", "Sw", "Pl"))
si_from_ecosite_guide("Boreal Plain Ecozone", "10", c("Pj", "Aw", "Sw"))
si_from_ecosite_guide("EcoSeries 30", "34", "Sw", details = TRUE)
```

The available classification units:

```{r}
unique(si_ecosite_guides[, c("jurisdiction", "unit")])
```

Soil moisture coding is not consistent between guides -- British Columbia,
Alberta and the 1996 Saskatchewan guide use the numeric 1-9 scale, while the
2010 Saskatchewan guide and the Manitoba guide use letter classes. The
published strings are carried through unchanged, so check `smr` and `snr`
against the guide named in the `guide` column before comparing across
jurisdictions.

## Worked example: filling gaps in a species list

A mixedwood plot in the Central Mixedwoods where only aspen carried a
measurable site index.

```{r}
plot_species <- data.frame(
  species = c("Aw", "Sw", "Sb", "Pl"),
  si_bh   = c(21.0, NA, NA, NA),
  stringsAsFactors = FALSE
)

# 1. edatope first, where the study covers the species
plot_species$si_edatope <- si_from_edatope("CM", 5, "C", plot_species$species) |>
  suppressWarnings()

# 2. species conversion from the measured aspen value as a cross-check
plot_species$si_converted <- ab_si_to_si("Aw", 21.0, plot_species$species) |>
  suppressWarnings()

# 3. resolve: measured, then edatope, then conversion
plot_species$si_final <- ifelse(
  !is.na(plot_species$si_bh), plot_species$si_bh,
  ifelse(!is.na(plot_species$si_edatope), plot_species$si_edatope,
         plot_species$si_converted)
)

plot_species

# GYPSY needs total-age site index and total age
plot_species$si_total <- ab_si_bh_to_total(plot_species$species,
                                           plot_species$si_final)
plot_species$topht_10 <- ab_si_to_height(plot_species$species,
                                         plot_species$si_total, total_age = 10)
plot_species[, c("species", "si_final", "si_total", "topht_10")]
```

Black spruce has no edatope entry for this cell and no conversion equation from
aspen, so it stays `NA` rather than silently borrowing another species' value.

## References

Bjelanovic, I., and Comeau, P.G. 2019. Estimating site index using ecosite and
edatope in Alberta and Saskatchewan. MGM Research Note #2019-1. University of
Alberta, Edmonton, Alberta.

Bjelanovic, I., and Comeau, P.G. 2019. Species SI conversion equations for
Alberta and Saskatchewan. MGM Research Note #2019-2. University of Alberta,
Edmonton, Alberta.

Comeau, P.G. 2020. Estimating site index using ecosite guides for Western
Canada. MGM Research Note #2020-1. University of Alberta, Edmonton, Alberta.

Huang, S., Meng, S.X., and Yang, Y. 2009. A Growth and Yield Projection System
(GYPSY) for Natural and Post-harvest Stands in Alberta. Alberta Sustainable
Resource Development Tech. Rep. T/216.
