---
title: "Site Tree Workflow: Full API Walkthrough"
author: "CanWestSiteIndexCFS Contributors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Site Tree Workflow: Full API Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
library(CanWestSiteIndexCFS)
```

## Overview

This vignette walks through a complete site productivity estimation from raw
field data to downstream growth model inputs, exercising every public function
in the package. The five phases follow the function map in the package README.

**Scenario:** Six site trees from two mixed-species stands — interior
Douglas-fir (entered as `"FD"`) and interior spruce (entered as `"Sw"`).
Ages were measured at different reference points: some trees were bored at
the stump (total age), others at breast height (BH age).

> **Site tree requirement.** Height and age must come from a *suitable site
> tree* — a vigorous dominant or co-dominant tree with a full, undamaged crown
> and a straight stem free of disease, suppression, or open-grown influence.
> See `?CanWestSiteIndexCFS` for the full BC definition. This package does not select
> site trees from a general treelist.

---

## Input data

```{r data}
trees <- data.frame(
  tree_id     = 1:6,
  species_raw = c("FD",    "FD",  "FD",    "Sw",  "Sw",    "Sw"),
  height_m    = c(22.1,    19.8,  24.3,    18.5,  20.2,    17.9),
  age         = c(52L,     48L,   55L,     45L,   50L,     42L),
  age_source  = c("stump", "BH",  "stump", "BH",  "stump", "BH"),
  fiz         = "H",
  stringsAsFactors = FALSE
)
trees
```

`fiz = "H"` (Prince George Forest District) places all trees in the interior.
This matters in the next step: the code `"FD"` is ambiguous between coastal
and interior Douglas-fir, and the FIZ resolves it.

---

## ① Species Resolution

### Canonical species codes

`species_code()` normalizes any input — any capitalization, with or without
the coast/interior suffix — to the canonical form used internally by SiteTools.

```{r sp-code}
trees$sp_code <- species_code(trees$species_raw, fiz = trees$fiz)
trees[, c("tree_id", "species_raw", "sp_code")]
```

`"FD"` in FIZ `"H"` resolves to `"Fdi"` (interior Douglas-fir). Without a
FIZ, `"FD"` alone is ambiguous and `species_to_sp_index()` will throw an
error, which is intentional — silently defaulting to one coast/interior form
would be a source of hard-to-detect mistakes.

### Full species names

```{r sp-name}
unique_sp <- unique(trees$sp_code)
data.frame(
  sp_code = unique_sp,
  sp_name = species_name(unique_sp)
)
```

### Geographic occurrence

```{r sp-location}
species_location(unique_sp)
```

The `coast`, `interior`, and `common` columns are decoded from the SiteTools
bit-field. A species flagged for both coast and interior has curves registered
for both regions.

### Integer species index

All conversion functions ultimately work with an integer species index.
`species_to_sp_index()` is the FIZ-aware resolver; it is called internally
by every other function but is exposed when you need the raw index directly.

```{r sp-index}
species_to_sp_index(unique_sp, fiz = "H")
```

---

## ② Curve Inspection (optional)

Each species has one or more registered SI curves. `curve_options()` returns
a data frame listing every curve together with its index and a flag marking
the package default.

```{r curve-options}
curve_options("Fdi")
curve_options("Sw")
```

For most workflows the default curve is appropriate. When you need to
select a specific curve — for example, to match results from another tool
or to use a growth-intercept curve — inspect the full curve metadata first.

```{r curve-meta}
fdi_opts   <- curve_options("Fdi")
fdi_default_cu <- fdi_opts$curve_index[fdi_opts$is_default]

data.frame(
  curve_index = fdi_default_cu,
  name        = curve_name(cu_index   = fdi_default_cu),
  notes       = curve_notes(cu_index  = fdi_default_cu),
  source      = curve_source(cu_index = fdi_default_cu)
)
```

`curve_notes()` often contains important usage caveats (for example, age
range restrictions). `curve_source()` gives the full bibliographic citation,
which is useful for audit trails and reporting.

For workflows where the default curve depends on establishment type:

```{r curve-estab}
# Default curve by establishment type
default_curve_estab("Fdi", estab = 1)
default_curve_estab("Sw",  estab = 1)
```

`default_gi_curve()` returns the growth-intercept (GI) curve for each
species. GI curves are calibrated for younger trees and should be used
when breast-height age is **≤ 50 years**; for older trees the standard
SI curve is more appropriate. This is the same rule applied in BC
government compilers (e.g., FAIBCompiler).

```{r curve-gi}
default_gi_curve("Fdi")
default_gi_curve("Sw")
```

---

## ③ Age Preparation

`ht_age_to_si()` requires **breast-height age** (`age_type = 1`). Trees bored
at the stump have total age and must be converted first using `age_to_age()`.

```{r age-prep}
is_stump <- trees$age_source == "stump"

# Start with each tree's measured age; overwrite for stump-bored trees
trees$bh_age <- trees$age

trees$bh_age[is_stump] <- mapply(
  function(age, sp) {
    age_to_age(
      age1      = age,
      age1_type = 0L,   # 0 = total age (stump bore)
      age2_type = 1L,   # 1 = breast-height age
      species   = sp
    )
  },
  trees$age[is_stump],
  trees$sp_code[is_stump]
)

trees[, c("tree_id", "sp_code", "age", "age_source", "bh_age")]
```

**How y2bh is handled:** `age_to_age()` needs years-to-breast-height (y2bh)
to perform the conversion. When `y2bh` is not supplied it calls `si_to_y2bh()`
internally using a default SI of 30 m. If you already have an SI estimate —
for example from a prior measurement — you can supply `y2bh` explicitly for
slightly greater accuracy:

```{r age-prep-explicit-y2bh, eval=FALSE}
# Supplying y2bh explicitly (more accurate when SI is already known)
age_to_age(age1 = 52, age1_type = 0L, age2_type = 1L,
           species = "Fdi", y2bh = si_to_y2bh(species = "Fdi", site_index = 23))
```

---

## ④ Site Index Estimation

The appropriate curve depends on tree age:

- **BH age > 50**: use the default SI curve (reference age 50 BH)
- **BH age ≤ 50**: use the growth-intercept (GI) curve, which is more
  reliable for younger trees

First resolve both curves for each species:

```{r curve-resolve}
sp_curves <- do.call(rbind, lapply(unique(trees$sp_code), function(sp) {
  opts <- curve_options(sp)
  data.frame(
    sp_code  = sp,
    si_curve = opts$curve_index[opts$is_default],
    gi_curve = default_gi_curve(sp),
    stringsAsFactors = FALSE
  )
}))

# Attach to trees by lookup (avoids row reordering)
trees$si_curve <- sp_curves$si_curve[match(trees$sp_code, sp_curves$sp_code)]
trees$gi_curve <- sp_curves$gi_curve[match(trees$sp_code, sp_curves$sp_code)]
sp_curves
```

Now compute SI using the age-appropriate curve:

```{r si-calc}
trees$si <- mapply(
  function(age, ht, si_cu, gi_cu) {
    cu <- if (age <= 50L) gi_cu else si_cu
    ht_age_to_si(cu_index = cu, age = age, age_type = 1L, height = ht)
  },
  trees$bh_age,
  trees$height_m,
  trees$si_curve,
  trees$gi_curve
)

# ht_age_to_si() returns the raw SIndex numeric result.
# Negative values are SIndex error codes, not site index values.
# Common causes: height at or below BH (<=1.3 m), age of 0,
# or inputs outside the curve's valid range.
n_err <- sum(trees$si < 0, na.rm = TRUE)
if (n_err > 0) {
  warning(n_err, " tree(s) returned SIndex error codes -- set to NA.")
  trees$si[trees$si < 0] <- NA_real_
}

trees[, c("tree_id", "sp_code", "bh_age", "si_curve", "gi_curve", "height_m", "si")]
```

### Alternative: site class

When stand age is under approximately 30 years, SI from height and age is
unreliable. `site_class_to_index()` provides an alternative entry point using
a categorical site quality rating (G = good, M = medium, P = poor, L = low):

```{r site-class}
site_class_to_index("Fdi", site_class = "G")
site_class_to_index("Sw",  site_class = "M")
```

---

## Summarize by Species

Aggregate to a single SI estimate per species — the usual input for
stand-level growth modelling.

```{r si-summary}
si_summary <- aggregate(si ~ sp_code, data = trees, FUN = mean)
si_summary
```

### PSP vs non-PSP aggregation

The approach above — compute SI per tree, then average — matches the
**non-PSP** pattern used in BC government compilers (e.g., FAIBCompiler's
`siteAgeSummary()`).

**PSP data follows a different order.** In PSP programs, heights and ages
are averaged across suitable site trees first, and then SI is computed once
from those averaged values:

```{r si-summary-psp, eval=FALSE}
# PSP pattern: average height and age first, then compute SI once
psp_by_sp <- aggregate(cbind(height_m, bh_age) ~ sp_code, data = trees, FUN = mean)
psp_by_sp$si <- mapply(
  function(sp, age, ht, si_cu, gi_cu) {
    cu <- if (age <= 50L) gi_cu else si_cu
    ht_age_to_si(cu_index = cu, age = age, age_type = 1L, height = ht)
  },
  psp_by_sp$sp_code,
  psp_by_sp$bh_age,
  psp_by_sp$height_m,
  sp_curves$si_curve[match(psp_by_sp$sp_code, sp_curves$sp_code)],
  sp_curves$gi_curve[match(psp_by_sp$sp_code, sp_curves$sp_code)]
)
psp_by_sp
```

These two approaches produce different SI values because
`mean(SI(h_i, a_i))` ≠ `SI(mean(h_i), mean(a_i))`. Use the pattern
that matches how your source data was collected and how the downstream
compiler expects SI to be supplied.

Note also that PSP compilers (e.g., FAIBCompiler's `siteAgeSummary_PSP()`)
apply a looser GI validity threshold: both GI and standard SI curves are
computed, GI is preferred when `SI_GI > 0 & AGE_BH > 10`, with standard SI
as fallback. The `age ≤ 50` rule shown earlier in this vignette reflects the
non-PSP path.

---

## ⑤ Downstream Uses

### Project height at a future age

`si_age_to_ht()` answers: *given this site index, how tall will dominant trees
be at a target breast-height age?*

```{r si-age-to-ht}
si_summary$ht_at_bh100 <- mapply(
  function(sp, si) si_age_to_ht(age = 100L, age_type = 1L, site_index = si, species = sp),
  si_summary$sp_code,
  si_summary$si
)
si_summary
```

### Estimate age at a target height

`si_ht_to_age()` inverts the relationship — given SI and a target dominant
height, it returns the expected breast-height age:

```{r si-ht-to-age}
data.frame(
  sp_code        = si_summary$sp_code,
  si             = round(si_summary$si, 1),
  bh_age_at_30m  = mapply(
    function(sp, si) si_ht_to_age(site_height = 30, age_type = 1L, site_index = si, species = sp),
    si_summary$sp_code,
    si_summary$si
  )
)
```

### Years to breast height

`si_to_y2bh()` and `si_to_y2bh05()` return the expected number of years a
seedling takes to reach breast height (1.3 m), given species and site index.
This value is used internally by `age_to_age()` but is also needed explicitly
when converting between total and BH ages in plantation records, or when
building y2bh columns for growth model input tables.

```{r y2bh}
data.frame(
  sp_code    = si_summary$sp_code,
  si         = round(si_summary$si, 1),
  y2bh       = mapply(si_to_y2bh,   species = si_summary$sp_code, site_index = si_summary$si),
  y2bh_round = mapply(si_to_y2bh05, species = si_summary$sp_code, site_index = si_summary$si)
)
```

`si_to_y2bh05()` rounds to the nearest 0.5-year step, which is the convention
used by some growth models and SiteTools itself.

### Convert SI between species

`si_to_si()` translates a site index value from one species' curve to another
using the SiteTools species-conversion table. SI values on different curves
are not directly comparable — each curve was calibrated to data from that
species only — so `si_to_si()` is needed before combining or comparing site
index across species.

```{r si-to-si}
si_fdi <- si_summary$si[si_summary$sp_code == "Fdi"]
si_sw  <- si_summary$si[si_summary$sp_code == "Sw"]

data.frame(
  conversion = c("Fdi  ->  Sw", "Sw   ->  Fdi"),
  result_m   = c(
    si_to_si("Fdi", si_fdi, "Sw"),
    si_to_si("Sw",  si_sw,  "Fdi")
  )
)
```

---

## Summary

| Phase | Functions used |
|---|---|
| ① Species resolution | `species_code()`, `species_name()`, `species_location()`, `species_to_sp_index()` |
| ② Curve inspection | `curve_options()`, `curve_name()`, `curve_notes()`, `curve_source()`, `default_curve_estab()`, `default_gi_curve()` |
| ③ Age preparation | `age_to_age()`, `si_to_y2bh()` (internal) |
| ④ SI estimation | `ht_age_to_si()`, `site_class_to_index()` |
| ⑤ Downstream uses | `si_age_to_ht()`, `si_ht_to_age()`, `si_to_si()`, `si_to_y2bh()`, `si_to_y2bh05()` |
