---
title: "CanWestSiteIndexCFS Workflow Integration Examples"
author: "CanWestSiteIndexCFS Contributors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CanWestSiteIndexCFS Workflow Integration Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(CanWestSiteIndexCFS)
```

## Overview

This vignette shows two practical integration patterns:

1. Permanent Sample Plot (PSP) site productivity estimation.
2. Treelist preparation for growth and yield model inputs.

The examples use generic sample datasets included in the package under `inst/examples`.

## 1) PSP Workflow Example

Typical PSP programs maintain dominant height and age observations by plot and measurement year.
Using those fields, CanWestSiteIndexCFS can derive site index and years to breast height.

> **Site tree assumption.** The `dom_height_m` and `bh_age` fields in this example represent
> measurements taken from pre-designated **site trees** — vigorous dominant or co-dominant trees
> selected in the field according to BC's site tree criteria (see `?CanWestSiteIndexCFS` for the full
> definition). This package does **not** select site trees from a general treelist. It assumes
> that suitable trees have already been identified and that the supplied height and age values
> reflect those trees. Passing height or age values from non-site trees (e.g., suppressed,
> damaged, or open-grown trees) will produce site index estimates that are not ecologically
> meaningful.

```{r psp-load}
psp_path <- system.file("examples", "psp_sample.csv", package = "CanWestSiteIndexCFS")
psp <- read.csv(psp_path, stringsAsFactors = FALSE)
psp
```

```{r psp-calc}
# Resolve the default curve for each species — this gives us both the curve
# index (cu_index) and the equation name used for the site index calculation.
psp$cu_index <- mapply(
  function(sp) {
    df <- curve_options(sp)
    df$curve_index[df$is_default]
  },
  psp$species
)

psp$curve_name <- vapply(
  psp$cu_index,
  function(cu) curve_name(cu_index = cu),
  character(1)
)

psp$curve_source <- vapply(
  psp$cu_index,
  function(cu) curve_source(cu_index = cu),
  character(1)
)

# Now compute SI and y2bh using the resolved curve index directly.
psp$site_index_m <- mapply(
  function(cu, age, age_type, h, est_type) {
    ht_age_to_si(cu_index = cu, age = age, age_type = age_type, height = h, si_est_type = est_type)
  },
  psp$cu_index, psp$bh_age, psp$age_type, psp$dom_height_m, psp$si_est_type
)

psp$y2bh_years <- mapply(
  function(cu, si) si_to_y2bh(cu_index = cu, site_index = si),
  psp$cu_index, psp$site_index_m
)

psp
```

```{r psp-summary}
aggregate(
  psp[, c("site_index_m", "y2bh_years")],
  by = list(plot_id = psp$plot_id),
  FUN = mean
)
```

### Why This Helps

- **Site index (`site_index_m`)** gives a comparable productivity metric among plots.
- **Years to breast height (`y2bh_years`)** supports age conversions and growth model setup.
- **`cu_index` and `curve_name`** document exactly which equation was applied — important for auditability and reproducing results with other tools such as SiteTools.

## 2) Treelist Workflow for Growth and Yield Inputs

A common workflow is to estimate stand-level productivity from dominant age/height,
then attach those attributes to each tree record used by a growth model.

> **Site tree assumption.** The `dom_height_m` and `dom_bh_age` fields in `stand_inputs_sample.csv`
> represent stand-level dominant height and age derived from suitable site trees. When working from
> a full treelist rather than pre-compiled stand summaries, dominant height must be derived
> externally before calling `ht_age_to_si()` — for example, by selecting the tallest live
> dominant or co-dominant trees (crown class D or C) that are undamaged, have intact tops, and
> are free from suppression or other influences on height-growth potential. See `?CanWestSiteIndexCFS`
> for BC's full site tree definition.

```{r treelist-load}
stands_path <- system.file("examples", "stand_inputs_sample.csv", package = "CanWestSiteIndexCFS")
trees_path <- system.file("examples", "treelist_sample.csv", package = "CanWestSiteIndexCFS")

stands <- read.csv(stands_path, stringsAsFactors = FALSE)
trees <- read.csv(trees_path, stringsAsFactors = FALSE)

stands
trees
```

```{r treelist-calc}
stands$site_index_m <- mapply(
  function(sp, age, h) ht_age_to_si(age = age, age_type = 1, height = h, species = sp),
  stands$dom_species, stands$dom_bh_age, stands$dom_height_m
)

stands$y2bh_years <- mapply(
  function(sp, si) si_to_y2bh(site_index = si, species = sp),
  stands$dom_species, stands$site_index_m
)

stands$proj_height_m <- mapply(
  function(sp, age, si, y2bh) si_age_to_ht(age = age, age_type = 1, site_index = si, y2bh = y2bh, species = sp),
  stands$dom_species, stands$projection_age, stands$site_index_m, stands$y2bh_years
)

stands
```

```{r treelist-join}
model_treelist <- merge(
  trees,
  stands[, c("stand_id", "site_index_m", "y2bh_years", "projection_age", "proj_height_m")],
  by = "stand_id",
  all.x = TRUE,
  sort = FALSE
)

model_treelist
```

### Why This Helps

- The treelist receives consistent stand productivity fields.
- Model inputs can include both current tree attributes and site productivity descriptors.

## Run As Scripts

You can run the equivalent script versions directly:

```r
source(system.file("examples", "psp_workflow_example.R", package = "CanWestSiteIndexCFS"))
source(system.file("examples", "treelist_workflow_example.R", package = "CanWestSiteIndexCFS"))
```

